Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-28 doc/report/report.tex
report

\subsection{Swedish}
Swedish is a North-Germanic language spoken,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
A list of things:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is Subject-Verb-Object, but any syntactic category can be fronted.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} are being emphasized.
\enumsentence{
\shortex{4}
{Jag & ser & dig & inte.}
{I &  see & you & not.}
{`I  don't see you'.}} \label{ex:swedish-svo}
\enumsentence{
\shortexnt{4}
{Dig  & ser &  jag & inte.}
{You& I& see & not.}}\label{ex:swedish-ovs} 
\enumsentence{
\shortexnt{4}
{Inte & ser& jag &dig.}
{Not&  see& I&   you.}} \label{ex:swedish-avso}

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & jag & dig & inte?}
{Saw  & I&  you& not?}
{`Didn't I see you?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

\subsubsection*{Passive voice}
Passive voice is often used in Swedish, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb.
\enumsentence{
\shortex{5}{Han & jagade\textbf{s} & av & ett & lejon.}
        {He & hunted+\textbf{s} & by & a & lion.}
        {`He was hunted by a lion.'}}
cf.
\enumsentence{
\shortex{5}{Ett & lejon & jagade & honom.}
        {A &lion &hunted &him.}
        {`A lion was hunting him.'}}

@@

\subsubsection*{Reflexive pronouns}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog sig. & Han såg sitt barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog honom. & Han såg hans barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses.
Covers sentences like
han har inte ätit två gula äpplen idag
word order, agreement, tempus, adverbs, adjectivs...
and relatives
Also info in the Extra module, where more language specific constructions
can be found.
jag ser dig
dig ser jag
dig har jag inte sett
jag har inte sett dig.
how the parse tree is generated, the finite verb, the subject, the rest. usw.
han såg katten han tyckte om
för vem måste jag akta mig
vem måste jag akta mig för

>>>>

\subsection{Swedish}
Swedish\cite{SAG:intro} is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite{H&H:1027}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{Saw  & you&  me& not?}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

\subsubsection*{Passive voice}
Passive voice is often used in Swedish, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb.
\enumsentence{Passive \hspace{39mm} Active\\
\shortexm{11}{Han & jagade\textbf{s} & av & ett & lejon.}
        {He & hunted+\textbf{s} & by & a & lion.}
        {`He was hunted by a lion.'}
{Ett & lejon & jagade & honom.}
        {A &lion &hunted &him.}
        {`A lion was hunting him.'}}

%cf.
%\enumsentence{
%\shortex{5}{Ett & lejon & jagade & honom.}
%        {A &lion &hunted &him.}
%        {`A lion was hunting him.'}}

@@

\subsubsection*{Reflexive pronouns}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
Sentence \ref{gfSwe:parseable} produces the parse in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag}
{He& has& not& eaten& the& yellow& apples& today}} %correct!!!
\label{gfSwe:parseable}
\begin{figure}[h]
\includegraphics[width=60mm]{gfTree.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit..} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. In the code, this is implemented by using a record.
The linearization rule for questions -- clauses with inverted order -- picks the 
finite form: \\
\verb|table { Inv => verb.fin ++ subj ++ neg ++ verb.inf ++ compl ; ... | \\
Relative clauses are also covered: 
\enumsentence{Hon ser pojken som sover\\
              She sees the boy who sleeps}

In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
The \verb|Extra| module included functions for topologication, although all
of them were not implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
Check wich were here first:
jag ser dig
dig ser jag
dig har jag inte sett
jag har inte sett dig.
how the parse tree is generated, the finite verb, the subject, the rest. usw.
han såg katten han tyckte om
för vem måste jag akta mig
vem måste jag akta mig för
